---
title: "Comparing groups"
subtitle: "REFERENCE SHEET"
format: 
  pdf:
    toc: true
    toc-depth: 5
    include-in-header: 
      text: |
        \usepackage{fvextra}
        \DefineVerbatimEnvironment{Highlighting}{Verbatim}{
          commandchars=\\\{\},
          breaklines, breaknonspaceingroup, breakanywhere
        }
execute:
  message: false
  warning: false
  echo: false
  fig-width: 8
  fig-height: 4
---

```{r}
#| label: packages

library(tidyverse)
library(flextable)
library(ggridges)
```

```{r}
#| label: data

# reading in data from "raw" csv stored in repo: https://github.com/ENVS-193DD/datasets
wq <- read_csv("https://raw.githubusercontent.com/ENVS-193DD/datasets/refs/heads/main/water-quality/wq_clean.csv") |> 
  # reordering factor levels
  mutate(site_name = fct_relevel(site_name,
                                 "Phelps Bridge", "East Channel", "Venoco Bridge"))
```

```{r}
#| label: plot-aesthetics

# colors to use in all figures
site_cols <- c(
  "East Channel" = "orange2",
  "Phelps Bridge" = "tomato3",
  "Venoco Bridge" = "turquoise4"
)

# setting common theme elements
theme_set(
  theme_classic() +
    theme(text = element_text(size = 12),
          plot.title.position = "plot",
          legend.position = "none") 
)

```


# 1. Math

## Mean

$$
\bar{y} = \frac{1}{n}\sum_{i = 1}^ny_i
$$

$\bar{y}$ is the sample mean, which is calculated as the sum of all observations $y_i$ divided by the number of observations ($n$).

## Median

The median is determined by finding the middle value. If $n$ is odd, the median is the middle value in the dataset. If $n$ is even, the median is the mean of the two middle values.

## Standard deviation

$$
s = \sqrt{\frac{\sum(y_i - \bar{y})^2}{n - 1}}
$$

$s$ is the sample standard deviation, which is calculated as the square root of the sum of the squared difference between an observation $y_i$ and the sample mean $\bar{y}$ divided by the number of observations $n$ minus 1.

## Variance

**Standard Deviation Squared**
- We used SD for description of spread because variance would be unit^2, so that's why we typically interpret it using SD.

$$
s^2 = \frac{\sum(y_i - \bar{y})^2}{n - 1}
$$

Variance is calculated as the standard deviation squared. Note that the units of variance are the units squared (e.g. g^2^, m^2^, ppt^2^).

Note that if you're describing spread in writing, it's usually best to go with standard deviation since the units are the same as the variable.

# 2. Calculating central tendency and spread

Prior to visualizing, it's sometimes helpful to actually calculate the metrics of central tendency (mean and median) and spread (standard deviation and variance) along with the sample size.

```{r}
#| label: wq-summary

# creating new object from water quality
salinity_summary <- wq |> 
  # group by site
  group_by(site_name) |> 
  # calculate mean, median, standard deviation, variance, number of obs
  summarize(mean_sal = mean(mean_salinity_ppt),
            median_sal = median(mean_salinity_ppt),
            sd_sal = sd(mean_salinity_ppt),
            var_sal = var(mean_salinity_ppt),
            n = length(mean_salinity_ppt)) |> 
  # ungroup data frame
  ungroup()

# use flextable to create table
flextable(salinity_summary) |> 
  # change column names in table to be readable
  set_header_labels(
    site_name = "Site",
    mean_sal = "Mean",
    median_sal = "Median",
    sd_sal = "Standard deviation",
    var_sal = "Variance",
    n = "Number of surveys"
  ) |> 
  # round any columns with numbers to one decimal point
  colformat_double(j = c("mean_sal", "median_sal", "sd_sal", "var_sal"), 
                   digits = 1) |> 
  # setting column width of table
  width(width = 1,
        unit = "in")
```

# 3. Visualizing distributions

The purpose of visualizing distributions is to represent the distribution of variable of interest.

Note that the interpretation of these plots is more or less the same: Most salinity measurements from Phelps Bridge are low. Most salinity measurements from East Channel and Venoco Bridge are below 40 ppt.

## Plots to start with

### Histograms

**Components**

- x-axis: _continuous_ variable of interest
- y-axis: counts
- Column height represents number of observations falling into a given bin range on the x-axis

**Pay attention to:** 

- the shape of the distribution
- the center of the distribution
- the peak(s) of the distribution

```{r}
#| label: histogram

# creating object for histogram
histogram <- ggplot(data = wq,
                    # x-axis: variable of interest
                    # fill geometries by site
                    mapping = aes(x = mean_salinity_ppt,
                                  fill = site_name)) +
  # histogram geometry
  geom_histogram(bins = 8,             # using 8 bins
                 color = "black") +    # coloring outline black
  # creating different panels for each site
  facet_wrap(~ site_name) +
  # getting rid of space between columns and axis, adding space at top
  scale_y_continuous(expand = expansion(mult = c(0, 0.025))) +
  # filling columns
  scale_fill_manual(values = site_cols) +
  # relabelling axes, adding title
  labs(x = "Salinity (ppt)",
       y = "Count",
       title = "Histograms")

# displaying histogram
histogram
```

- East and Venoco have **bimodal** distribution
- Phelps is **unimodal** with lots of observations at low values

Potential caption:

**Figure 1. Distributions of salinity differ across sites.** Distributions of salinity (parts per thousand) shown in histograms for each site (red: Phelps Bridge, orange: East Channel, and teal: Venoco Bridge). Bars represent counts of observations falling in each bin along the x-axis.

### Boxplots

**Components:**

- x-axis: categories (levels in a factor)
- y-axis: variable of interest
- middle line: median
- bottom of box: first quartile (25th percentile)
- top of box: third quartile (75th percentile)
- whiskers: 1.5 $\times$ interquartile range
- points: observation values outside 1.5 $\times$ interquartile range

**Pay attention to**
- 1. median position relative to y-axis
- 2. median position relative to each other

```{r}
#| label: boxplots

# creating object for boxplots
boxplot <- ggplot(data = wq,
                  # x-axis: site
                  # y-axis: salinity
                  # fill geometries by site
                  mapping = aes(x = site_name,
                                y = mean_salinity_ppt,
                                fill = site_name)) +
  # boxplot (and narrowing width)
  geom_boxplot(width = 0.5) +
  # filling geometries
  scale_fill_manual(values = site_cols) +
  # relabelling x- and y-axes, adding title
  labs(x = "Site",
       y = "Salinity (ppt)",
       title = "Boxplot")

# displaying object
boxplot
```

Potential captions:

**Figure 2. Median salinity differs across sites.** Boxplots representing median (middle line) salinity (parts per thousand) and 25th and 75th percentiles at each site (red: Phelps Bridge, orange: East Channel, and teal: Venoco Bridge). Whiskers represent observations within 1.5 $\times$ interquartile range, and points represent observations outside the range. 

### Strip charts

**Components:**

- x-axis: categories (levels in a factor)
- y-axis: variable of interest
- geometry: jittered points

**Pay attention to:**

- the clusters of points, if any
- position of points along the y-axis

```{r}
#| label: strip-chart

# creating new object for jitter plot
jitter <- ggplot(data = wq,
                 # x-axis: site
                 # y-axis: salinity
                 # color geometries by site
                 mapping = aes(x = site_name,
                               y = mean_salinity_ppt,
                               color = site_name)) +
  # jitter with adjustments
  geom_jitter(height = 0,     # no jitter along y-axis
              alpha = 0.5,    # transparent shapes
              width = 0.2,    # narrowing jitter
              shape = 21) +   # open circles
  # coloring 
  scale_color_manual(values = site_cols) +
  # relabelling x- and y-axes, adding title
  labs(x = "Site",
       y = "Salinity (ppt)",
       title = "Strip chart")

# displaying object
jitter
```


**Figure Clustering Observations**
- Clustering around 0 for all sites
- Could be some cluster around > 20
- Phelps have narrower range
- East Channel and Venoco Bridge have a larger range

Potential captions:

**Figure 3. Median salinity differs across sites.** Boxplots representing median (middle line) salinity (parts per thousand) and 25th and 75th percentiles at each site (red: Phelps Bridge, orange: East Channel, and teal: Venoco Bridge). Whiskers represent observations within 1.5 $\times$ interquartile range, and points represent observations outside the range. 

## Taking things a step further

### Density plots

**Components:**

- x-axis: continuous variable of interest
- y-axis: *probability density*
- smoothed function representing the shape of the distribution
- note: area under that smoothed function = 1

Curve:
- "kernel density estimation"

*Only really good for large dataset*

Log 1 + transformation:
- good to better see patterns
- **be careful when making figure interpretation from the transform plot**. (3 on the transformed plot would have to be transformed back to find the true value)
- All of your descriptions of plots with log+1 transform should be qualitative

**Pay attention to:**

- the shape of the **distribution**
- the center of the distribution
- the peak(s) of the distribution

```{r}
#| label: density-plots

# creating object for density
density <- ggplot(data = wq, 
                  # x-axis: variable of interest
                  # filling geometries by site name
                  mapping = aes(x = mean_salinity_ppt,
                                fill = site_name)) +
  # density plot
  geom_density() +
  # creating different panels for each site
  facet_wrap(~ site_name) +
  # getting rid of space between density function and axis, adding space at top
  scale_y_continuous(expand = expansion(mult = c(0, 0.025))) +
  # filling geometries
  scale_fill_manual(values = site_cols) +
  # relabelling axes, adding title
  labs(x = "Salinity (ppt)",
       y = "Density",
       title = "Density plots")

# displaying object
density

# creating object for density
density_log1 <- ggplot(data = wq, 
                       # x-axis: variable of interest
                       # filling geometries by site name
                       mapping = aes(x = log(mean_salinity_ppt + 1),
                                     fill = site_name)) +
  # density plot
  geom_density() +
  # creating different panels for each site
  facet_wrap(~ site_name) +
  # getting rid of space between density function and axis, adding space at top
  scale_y_continuous(expand = expansion(mult = c(0, 0.025))) +
  # filling geometries
  scale_fill_manual(values = site_cols) +
  # relabelling axes, adding title
  labs(x = "log+1 salinity (ppt)",
       y = "Density",
       title = "Density plots (log+1 transformed)")

# displaying object
density_log1
```

Potential captions:

**Figure 4. Distributions of salinity differ across sites.** Distributions of salinity (parts per thousand) shown in density plots for each site (red: Phelps Bridge, orange: East Channel, and teal: Venoco Bridge). 

**Figure 5. Distributions of salinity differ across sites.** Distributions of log+1 transformed salinity (parts per thousand) shown in density plots for each site (red: Phelps Bridge, orange: East Channel, and teal: Venoco Bridge). 

:::{.callout-note title="Use log+1 transformation with caution!"}
The transformation can help with visualization (i.e. make it easier to pick out patterns), but can be tricky to interpret/analyze. Proceed with caution and ask yourself whether you really _need_ the transformation to see patterns.
:::

### Violin charts

**Note:** violin charts are density plots that are mirrored.

**Components:**

- x-axis: categories (levels in a factor)
- y-axis: variable of interest
- geometry: violin

**Pay attention to:**

- wide or narrow parts of the violin
- position of wide or narrow parts along y-axis

**Notes**

- density plot mirrored across y-axis
- allows you to better see the distributions

```{r}
#| label: violin

# creating new object for violin
violin <- ggplot(data = wq,
                 # x-axis: site
                 # y-axis: salinity
                 # fill geometries by site
                 mapping = aes(x = site_name,
                               y = mean_salinity_ppt,
                               fill = site_name)) +
  # geometry for violin
  geom_violin() +
  # filling geometries
  scale_fill_manual(values = site_cols) +
  # relabelling x- and y-axes, adding title
  labs(x = "Site",
       y = "Salinity (ppt)",
       title = "Violin plot")

# displaying object
violin

# creating new object for violin
violin_log1 <- ggplot(data = wq,
                      # x-axis: site
                      # y-axis: salinity
                      # fill geometries by site
                      mapping = aes(x = site_name,
                                    y = log(mean_salinity_ppt + 1 ),
                                    fill = site_name)) +
  # geometry for violin
  geom_violin() +
  # filling geometries
  scale_fill_manual(values = site_cols) +
  # relabelling x- and y-axes, adding title
  labs(x = "Site",
       y = "log+1 salinity (ppt)",
       title = "Violin plot (log+1 transformed)")

# displaying object
violin_log1
```

Figure Interpretations:
- Phelps has more observations at low end
- East and Venoco have more observations at high end

Potential captions:

**Figure 6. Distributions of salinity differ across sites.** Distributions of salinity (parts per thousand) shown in violin plots for each site (red: Phelps Bridge, orange: East Channel, and teal: Venoco Bridge). 

**Figure 7. Distributions of salinity differ across sites.** Distributions of log+1 transformed salinity (parts per thousand) shown in violin plots for each site (red: Phelps Bridge, orange: East Channel, and teal: Venoco Bridge). 


:::{.callout-note title="Plot pitfalls"}

Density plots, violin plots, and boxplots are appropriate if you have many observations per group (or level in a factor). If you have few observations, those plots will either 

a) hide the fact that you have few observations (as in boxplots), 
b) suggest the existence of values where there aren't any (as in density or violin plots)
c) or some combination of both.

There's no hard and fast rule for the number of observations that are appropriate. However, the solution to these problems is almost always to _show the underlying data_ in a jitter or something similar.

:::

# 4. Visualizing central tendency and spread

## Medians

Refer to the section on boxplots - these incorporate medians along with spread (quantiles).

## Means

**With standard deviation:**

```{r}
#| label: means-sd

# starting with jitter plot from previous chunk
jitter +
  # adding summary information
  geom_pointrange(data = salinity_summary,
                  mapping = aes(x = site_name,
                                y = mean_sal,
                                ymin = mean_sal - sd_sal,
                                ymax = mean_sal + sd_sal,
                                color = site_name),
                  size = 1,
                  linewidth = 1) +
  # adding different title
  labs(title = "Strip chart with means and standard deviation")
```

**With standard error:**

*Represents how precise you sample mean is*

Note that standard error is calculated as:

$$
\frac{s}{\sqrt{n}}
$$

where $s$ is the standard devation and $n$ is the number of observations in the sample.

```{r}
#| label: means-se

jitter +
  # different geometry to calculate standard error 
  stat_summary(geom = "pointrange",
               # argument to calculate standard error
               fun.data = mean_se,
               size = 1,
               linewidth = 1) +
  labs(title = "Strip chart with means and SE")
```

**With 95% confidence interval:**

You are 95% confident that the true mean salinity value falls within the 95% CI range

**If you include whiskers you need to include description of metric represented in the whisker**

```{r}
#| label: means-ci

jitter +
  # different geometry to calculate a 95% confidence interval
  stat_summary(geom = "pointrange",
               # argument to calculate 95% confidence interval
               fun.data = mean_cl_normal,
               size = 1,
               linewidth = 1) +
  labs(title = "Strip chart with means and 95% confidence interval")
```

:::{.callout-note title="Be explicit about what whiskers mean!"}

These plots all use the same geometry (`pointrange`) to represent the means and whiskers, but all three plots use whiskers to represent different values.

Whenever making a plot with means and whiskers, it's _very_ important to be explicit about what values those whiskers represent in the axis label and/or the caption.

:::

Potential caption:

**Figure 10. Mean salinity differs across sites.** Small open circles represent observations of salinity at each sampling point at each site (red: Phelps Bridge, orange: East Channel, and teal: Venoco Bridge). Large points represent means. Whiskers represent 95% confidence intervals around the mean.

# 5. Making new types of plots and combining geometries

There are lots of options to visualize variables to compare across groups. Uuse graph galleries like the [R Graph Gallery](https://r-graph-gallery.com/) or [From Data to Viz](https://www.data-to-viz.com/) to come up with new ideas to try.

You can also combine different geometries and plots to represent data in a new way.

This plot is called a **ridgeline plot** because the distributions resemble mountain ridges. Adding points to represent means and 95% confidence intervals adds a metric of central tendency to the plot.

```{r}
#| label: ridgeline-plot

# creating new object for ridgeline
ridges <- ggplot(data = wq,
                 # x-axis: salinity
                 # y-axis: site
                 # fill geometries by site
                 mapping = aes(x = mean_salinity_ppt,
                               y = site_name,
                               fill = site_name)) +
  # ridges (scale makes ridges shorter so that they don't overlap)
  geom_density_ridges(scale = 0.8) +
  # adding means and 95% confidence intervals
  stat_summary(geom = "pointrange",
               fun.data = mean_cl_normal,
               size = 0.75,
               linewidth = 1) +
  # filling geometries with color
  scale_fill_manual(values = site_cols) +
  # setting y-axis expansions
  scale_y_discrete(expand = expansion(mult = c(0.01, 0.15))) +
  # relabelling x- and y-axis
  labs(x = "Salinity (ppt)",
       y = "Site")

# displaying object
ridges
```

# 6. Saving figures

Sometimes it is useful to save figures as files, either to test out the dimensions or to use the figures elsewhere. 

You can save a figure as any file type (e.g. .jpg, .png, .svg, .pdf) using `ggsave()`.

```{r}
#| label: ggsave
#| eval: false

ggsave(
  # file path to store figure
  here::here("comparing-groups", "figures", "ridges.jpg"),
  # object name
  ridges,
  # figure dimensions
  width = 8,
  height = 6,
  units = "in",
  # figure resolutions (good to start with 300 dpi aka dots per inch)
  dpi = 300
)
```

# 7. Analyses to compare metrics of central tendency across groups

## Medians

if comparing two groups: **Mann Whitney U**

if comparing more than two groups: **Kruskal-Wallis**

## Means

if comparing two groups: **two-sample t-test**

if comparing more than two groups: **analysis of variance (ANOVA)**

## Counts

**chi-square**

# 8. Project work

Get with your group members and discuss:

- which of the plots above would be appropriate for your variables
- which - if any - of these plots map onto the potential figures you designed in your work plan (if they don't, modify one of your potential figures to include a comparison of groups)
- (if your main question is comparing a response variable across groups) decide which metric of central tendency and test you should be using (from part 6 of this sheet)

Then, 

- (if you haven't already) create a repo for your final project and make sure everyone is added as a collaborator
- create `code` and `data` folders
- create a file in your `code` folder for _drafts_ of your data visualization 
- work together to try 1-2 of the above plots
- work together to try 1-2 different plots from the [R Graph Gallery](https://r-graph-gallery.com/) or [From Data to Viz](https://www.data-to-viz.com/)

:::{.callout-note title="All final projects must include at least one figure in which you are comparing your response variable across groups."}

A figure in which you are comparing your response variable across groups might directly answer your question or might provide context for your question - either way, you should have at least _one_ figure comparing a metric of central tendency across groups.

:::

"execute:
  message: false
  warning: false
  echo: false
  fig-width: 8
  fig-height: 4"
  
- chooses to show code chunk + output or just output
